A comparison of single-cell trajectory inference methods: towards more accurate and robust tools

### 2018-03-05 analysis/paper/paper_latex.Rmd
commit latest version of rmd

```{r}
add_sfig(
  figure_file("methods_timeline.tmp.pdf", "4-method_characterisation"),
  "methods_over_time", 
  "New TI methods over time. a) Number of methods published or in preprint. b) Number of methods which can handle a particular type of trajectory."
)
```

<!-- §2: Plethora of new TI methods -->
A plethora of TI methods has been developed over the last years, and even more are being created every month (`r ref("sfig", "methods_over_time", "a")`). We gathered a list of `r sum(implementations$is_ti, na.rm=TRUE)` TI methods (`r ref("table", "implementations")`), which we believe to be the most complete list to date. It is perhaps surprising that almost all of them have a unique combination of characteristics, in terms of the required inputs (prior information), produced outputs (topology fixing and trajectory type) and methodology used.

```{r, results='asis' }
add_table(
  "implementations",
  read_rds(figure_file("implementations_table.rds", "4-method_characterisation")),
  "Overview of the trajectory inference methods included in this study, and several characteristics thereof. This table will be continuously updated at https://goo.gl/FnsTxG."
)
```
<!-- §3: Topology inference -->
One distinctive characteristic of TI methods is whether the topology of the trajectory is inferred computationally, or was fixed by design.
Early TI methods typically fixed the topology algorithmically (e.g. linear [@bendall_singlecelltrajectorydetection_2014; @schlitzer_identificationcdc1cdc2committed_2015; @shin_singlecellrnaseqwaterfall_2015; @campbell_bayesiangaussianprocess_2015] or bifurcating [@haghverdi_diffusionpseudotimerobustly_2016; @setty_wishboneidentifiesbifurcating_2016]), or through parameters provided by the user [@trapnell_dynamicsregulatorscell_2014; @matsumoto_scoupprobabilisticmodel_2016]. These methods therefore mainly focused on correctly ordering the cells along this fixed topology. 
Other methods attempt to infer the topology computationally, which increases the difficulty of the problem at hand, but allows these methods  to be broadly applicable on more use cases. Methods that perform topology inference are still in the minority, though current trends suggest this will soon change (`r ref("sfig", "methods_over_time", "c")`).  A particularly interesting development is presented in the AGA method [@wolf_graphabstractionreconciles_2017] which is the only TI method currently able to deal with disconnected graphs.

@@

```{r}
add_sfig(
  figure_file("implementation_components.tmp.pdf","4-method_characterisation"), 
  "implementation_components", 
  "TI methods frequently share steps."
)
```

@@

```{r}
add_sfig(
  figure_file('dataset_characterisation.tmp.pdf', '2-dataset_characterisation'),
  "dataset_characterisation",  
  "Characteristics of the real datasets used for the evaluation."
)
```

```{r}
add_sfig(
  figure_file('synthetic_data_generation.tmp.pdf', "2-dataset_characterisation/3-synthetic"),
  "synthetic_data_generation",
  "Synthetic data generation workflow."
)
```

@@

```{r}
add_sfig(
  figure_file('expanded_ti_evaluation_overview_v1.tmp.pdf', 'evaluation_overview'),
  "extended_evaluation_overview",
  "Extended overview", 
  15, 
  7.5
)
```

@@

```{r}
add_sfig(
  figure_file(
    "real_synthetic_comparison.tmp.pdf",
    "5-optimise_parameters/5-real_synthetic_comparison"
  ), 
  "real_synthetic_comparison", 
  "Comparison between performance on real and synthetic datasets across trajectory types."
)
```

<!-- §4: Compare metrics -->
As the different metrics were selected to assess the correctness of a trajectory in various approaches, it is expected to observe differences in rankings of TI methods amongst the different metrics. While we saw no direct link between the edge flip scores versus the correlation or RF MSE metric, methods that obtained a high correlation score also tended to obtain higher RF MSE scores (`r ref("sfig", "metric_comparison")`). In addition, the methods that were able to detect more complex trajectory types also obtained higher edge flip scores, in comparison to methods whose trajectory topology was fixed to a simple trajectory topology. We will explore this issue in more detail in a further section SEC_TRAJTYPE.

```{r}
add_sfig(
  figure_file('metrics_comparison.tmp.pdf', "5-optimise_parameters/7-metric_comparison"), 
  "metric_comparison", 
  "Comparison between the different metrics across trajectory types."
)
```

@@

```{r}
add_sfig(
  figure_file('trajtype_handle_comparison.tmp.pdf', "8-compare_topology"), 
  "trajtype_handle_comparison", 
  "ADD CAPTION"
)
```

```{r}
add_sfig(
  figure_file('linear_split_comparison.tmp.pdf', "8-compare_topology"), 
  "linear_split_comparison", 
  "Linear methods have a higher correlation score on non-linear datasets. Methods were split in linear methods (those which can either only detect cyclic or linear trajectories) and non-linear methods."
)
```

@@

```{r}
add_sfig(
  figure_file('topology_sensitivity.tmp.pdf', "8-compare_topology"), 
  "topology_sensitivity", 
  "ADD CAPTION"
)
```

@@

```{r}
add_sfig(
  figure_file('required_priors_performance_comparison.tmp.pdf', "10-prior_information"), 
  "required_priors_performance_comparison", 
  "Comparison of method performance."
)
```

@@

```{r}
add_sfig(
  figure_file("method_components_scores.tmp.pdf","6-developer_guidelines"), 
  "method_components_scores", 
  "Score of methods containing particular algorithmic components. NS: Not significant. *: "
)
```

```{r}
add_stable(
  read_rds(figure_file(
    "component_importance.rds",
    "6-developer_guidelines"
  ))[[params$table_format]], 
  "component_importance", 
  "Importance values for algorithmic components."
)
```

@@

```{r}
add_sfig(
  figure_file(
    "qc_performance_comparison.tmp.pdf",
    "5-optimise_parameters/6-qc_performance_comparison"
  ), 
  "qc_performance_comparison", 
  "Comparison between method qc (x-axis) and method performance (y-axis)."
)
```

@@

Our workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods [@schaffter_genenetweaversilicobenchmark_2011; @marbach_wisdom_2012] and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment (`r ref("sfig", "synthetic_data_generation")`). At every step, we took great care was taking to mimic real cellular regulatory networks as best as possible, while keeping the model simple and easily extendable. For every synthetic dataset, we used a random real dataset as a reference dataset (from those described earlier), making sure the number of variable genes and cells were similar.

@@

One of the main process involved in cellular dynamic processes is gene regulation, where regulatory cascades and feedback loops lead to progressive changes in expression and decision making. The exact way a cell chooses a certain path during its differentiation is still an active research field, although certain models have already emerged and been tested in vivo. One driver of bifurcations seem to be mutual antagonism, where genes [@xu_regulationbifurcatingcell_2015] strongly repress each other, forcing one of the two to become inactive [@graf_forcingcellschange_2009]. Such mutual antagonism can be modelled and simulated [@wang_quantifyingwaddingtonlandscape_2011; @ferrell_bistabilitybifurcationswaddington_2012]. Although such a two-gene model is simple and elegant, the reality is frequently more complex, with multiple genes (grouped into models) repressing each other [@yosef_dynamicregulatorynetwork_2013].

To simulate certain trajectory topologies, we therefore designed module networks in which the cells will follow a particular trajectory topology given certain parameters (`r ref("sfig", "modulenets")`). Two module networks generated linear trajectories (linear and linear long), two generated simple forks (bifurcating and converging), one generated a complex fork (trifurcating), one generated a rooted tree (consecutive bifurcating) and two generated trajectories which simple undirected graphs (bifurcating loop and bifurcating convergence).

From these module networks we generated gene regulatory networks in two steps: generation of the main regulatory network, and addition of extra target genes from real regulatory networks. For each dataset, we used the same number of genes as were differentially expressed in the real datasets. 5% of genes were assigned to be part of the main, and were randomly distributed among all modules (with at least one gene per module). We sampled edges between these individual genes (according to the module network) using an uniform distribution between 1 and the number of possible targets in each module. To add additional target genes to the network, we assigned every regulator from the network to a real regulator in a real network (from regulatory circuits [@marbach_tissuespecificregulatorycircuits_2016]), and extracted for every regulator a local network around it using personalized pagerank (with damping factor set to 0.1), as implemented in the `page_rank` function of the *igraph* package.

```{r}
add_sfig(
  read_rds(figure_file(
    "modulenets.rds",
     experiment_id = "2-dataset_characterisation/3-synthetic"
   )), 
  "modulenets", 
  "Module networks, milestone networks and the module dynamics for each of the different types of synthetic data", 
  15, 
  5
)
```

### Simulation of thermodynamics

To simulate the gene regulatory network, we used a system of differential equations similar to which has been standard in evaluations of network inference [@marbach_wisdom_2012]. In this model, changes in gene expression ($x_i$) and protein expression ($y_i$) are modeled using ordinary differential equations [@schaffter_genenetweaversilicobenchmark_2011] (ODEs):

@@

Because each cellular simulations follows the trajectory it its own speed, knowing the exact position of a cell within the trajectory topology is not straightforward. Furthermore, the speed by which simulated cells made a decision between two or more alternative paths was highly variable. To estimate a cell's position during a simulation within the trajectory topology, we therefore used the known progression of the modules, given in `r ref("sfig", "modulenets", "c")`,  as a backbone. We smoothed the expression in each simulation using a rolling mean with a window of 50 time steps, and then calculated the average module expression along the simulation. We used dynamic time warping, implemented in the dtw R package [@giorgino_computingvisualizingdynamic_2009; @tormene_matchingincompletetime_2009], with an open end to align a simulation to all possible module progressions, and then picked the alignment which minimised the normalised distance between the simulation and the backbone. In case of cyclical trajectory topologies, the number of possible milestones a backbone could progress through was limited to 20.

## Common normalisation pipeline
We used a standard single-cell RNA-seq preprocessing pipeline which applies parts of the scran and scater Bioconductor packages [@lun_stepbystepworkflowlowlevel_2016]. The advantages of this pipeline is that it works both with and without spike-ins, and includes a harsh cell filtering which looks at abnormalities in library sizes, mitochondrial gene expression, and number of genes expressed using median absolute deviations (set to 3). We required that a gene has to be expressed in at least 5% of the cells, and that it should have an average expression higher than 0.05. Furthermore, we used the pipeline to select the most highly variable genes, using a false discovery rate of 5% and a biological component higher than 0.5. As a final filter, we removed both all zero genes and cells until convergence.

## Evaluation metrics
We used three different metrics to assess the performance of each TI method (`r ref("fig", "evaluation_overview")`). The importance of using multiple metrics to compare complex models has been stated repeatedly [@norel_selfassessmenttrap_2011]. Indeed, each of our metrics looks at the correspondence between prediction and gold standard in a different way: the correlation between geodesic distances assesses the overall ordering, the RF prediction error assesses the cell neighbourhood, and the edge flip score assesses the milestone network topology (`r ref("sfig", "score_aspects")`).

```{r}
add_sfig({
  read_rds(figure_file(
    "score_aspects.rds", 
    experiment_id="3-score_rules/5-score_aspects"))
  }, 
  "score_aspects", 
  "Different scores assess different aspects of the correspondence between a prediction and gold standard trajectory.",
  15, 
  8
)
```

### Correlation between geodesic distances
The accuracy of cell ordering was assessed by first calculating the geodesic distances $d(c_i, c_j)$ between all cells in both the gold standard and the prediction. The geodesic distance is calculated by finding the shortest path between two cells through the milestone network, while taking into account the edge lengths as returned by most methods. To calculate this for every pair of cells, we first calculated the shortest path lengths $d(m_a, m_b)$ between all milestones using Dijkstra's algorithm, as implemented in the R *igraph* package. We also calculated $d(m_a, c_i)$ for every cell to its neighbouring milestones by calculating the earth's movers distance using the shortlist method [@gottschlich_shortlistmethodfast_2014], as implemented in the R *transport* package. Then we again used Dijkstra's algorithm to calculate the shortest path lengths between every pair of cells which minimises:

$$
d(c_i, c_j) = min_{a \in A, b \in B} d(m_a, c_i) + d(m_a, m_b) + d(m_b, c_j)
$$

where A and B contain all milestones adjacent to respectively $c_i$ and $c_j$.
The final score was then calculated the Spearman's rank correlation between all $d(c_i, c_j)$ in gold standard and prediction.

In some cases, methods are unable to place some cells on a trajectory, mainly because they classify these cells as noisy. In these cases, we add these cells to a new milestone and put this milestone at a distance of 5 times the maximal distance between two milestones within the network.

@@

Although the correlation between geodesic distances directly assesses the position of the cells in the trajectory, a bad correlation does not directly imply that similar cells were not grouped together by the method, as illustrated in `r ref("sfig", "score_aspects", c("b", "c"))`. For example, certain methods will inevitably reach an incorrect ordering because they cannot handle the correct trajectory type, but these methods could still correctly place similar cells next to each other. We therefore also included a metric which looks at the local neighbourhood of each cell, and assesses whether this neighbourhood can accurately predict the position of this cell in the gold standard. We used a Random Forest regression, implemented in the ranger package [@wright_rangerfastimplementation_2017] to separately predict milestone percentages of every cell in the gold standard, using the milestone percentages of these cells in the prediction as features. We then used the out-of-bag mean-squared error on these percentages to score each method's capability of predicting the correct neighbourhood of each cell.

@@

As a third independent score, we assessed the similarity between the milestone network topology. We first simplified each network, by merging consecutive linear edges into one edge, by adding new milestones within self loops such that $A \rightarrow A$ would be converted $A \rightarrow B \rightarrow C \rightarrow D$, by adding an intermediate node to linear networks. Because we are interested , the network was made undirected. Next, we define the edge flip score as the minimal number of edges which should be added or removed to convert one network into the other, divided by the total number of edges in both networks. This problem is equivalent to the maximum common edge subgraph problem, a known NP-hard problem with no scalable solutions [@bahiense_maximumcommonedge_2012]. We implemented a branch & bound approach for this problem, by first enumerating all possible edge additions and removals with the minimal number of edges (the edge difference between the two networks) and if none of the new networks was isomorphic, try out all solutions with an additional two edge changes. To further limit the search space, we make sure the degree distributions between the two networks are similar, before assessing whether the two networks are isomorphic using the BLISS algorithm [@junttila_engineeringefficientcanonical_2007], as implemented in the R *igraph* package.

```{r}
add_sfig(
  figure_file("edge_flip_comparison.tmp.pdf", experiment_id="3-score_rules/4-edge_flip"), 
  "edge_flip_comparison", 
  paste0("Edge flip score values for common trajectory topologies, defined in ", ref("sfig", "modulenets")), 
  15, 
  5
)
```

### Aggregation of scores
```{r}
add_sfig(
  figure_file("metric_aggregation_overview_v1.tmp.pdf", experiment_id="evaluation_overview"), 
  "aggregation_of_results", 
  "Aggregation methodology of metric scores.", 
  15, 
  5
)
```

@@

### Pipeline
<!-- Look at dyneval methods to write this paragraph -->

### Method execution
We ran each method in a controlled environment, where we limited the memory usage (max. 32GB) and total run time (max 24h). Moreover, we disabled the function of the R `set.seed` function, to avoid that methods false make their output deterministic.

To assess the effect of prior information on the performance of a method, we compared the performance of methods which do or do not require the prior information using a two-tailed Mann–Whitney U test. P-values were controlled for multiple testing using Benjamini-Hochberg correction.

>>>>

```{r}
add_sfig(
  figure_file("methods_timeline.tmp.pdf", "4-method_characterisation"),
  "methods_over_time", 
  "New TI methods over time. a) Number of methods published or in preprint. b) Number of methods fixing the trajectory topology, either by design (fixed) or through user parameters (parameter). c) Number of methods which can handle a particular type of trajectory."
)
```

```{r, results='asis' }
add_table(
  "implementations",
  read_rds(figure_file("implementations_table.rds", "4-method_characterisation")),
  "Overview of the trajectory inference methods included in this study, and several characteristics thereof. This table will be continuously updated [online](https://goo.gl/FnsTxG)."
)
```

<!-- §2: Plethora of new TI methods -->
A plethora of TI methods has been developed over the last years, and even more are being created every month (`r ref("sfig", "methods_over_time", "a")`). It is perhaps surprising that of the `r sum(implementations$is_ti, na.rm=TRUE)` methods in existence today, almost all methods have a unique combination of characteristics (`r ref("table", "implementations")`), in terms of the required inputs (prior information), produced outputs (topology fixing and trajectory type) and methodology used (not shown).
<!-- §3: Topology inference -->
One distinctive characteristic of TI methods is whether the topology of the trajectory is inferred computationally, or was fixed by design.
Early TI methods typically fixed the topology algorithmically (e.g. linear [@bendall_singlecelltrajectorydetection_2014; @schlitzer_identificationcdc1cdc2committed_2015; @shin_singlecellrnaseqwaterfall_2015; @campbell_bayesiangaussianprocess_2015] or bifurcating [@haghverdi_diffusionpseudotimerobustly_2016; @setty_wishboneidentifiesbifurcating_2016]), or through parameters provided by the user [@trapnell_dynamicsregulatorscell_2014; @matsumoto_scoupprobabilisticmodel_2016]. These methods therefore mainly focused on correctly ordering the cells along this fixed topology. 
Other methods attempt to infer the topology computationally, which increases the difficulty of the problem at hand, but allows these methods  to be broadly applicable on more use cases. Methods that perform topology inference are still in the minority, though current trends suggest this will soon change (`r ref("sfig", "methods_over_time", "c")`).  A particularly interesting development is presented in the AGA method [@wolf_graphabstractionreconciles_2017] which is the only TI method currently able to deal with disconnected graphs.

@@

```{r}
add_sfig(
  figure_file("implementation_components.tmp.pdf","4-method_characterisation"), 
  "implementation_components", 
  "Common algorithmic components shared by different TI methods. Most components can be categorised into 4 categories: clustering (pink), dimensionality reduction (orange), generative models (green) and graph building (blue). a) Wordcloud of the common components. b) Number of times a component was shared between methods. c) Number of methods containing a particular category."
)
```

@@

```{r}
add_sfig(
  figure_file('dataset_characterisation.tmp.pdf', '2-dataset_characterisation'),
  "dataset_characterisation",  
  "Characteristics of the real datasets used for the evaluation. Top: distributions of number of cells and genes (after filtering) and the date these datasets were published. Bottom: Diversity of technologies which were used to generate the dataset, the organism, whether we extracted a silver or a gold standard from the data, the type of dynamic process and the trajectory type present in the data."
)
```

```{r}
add_sfig(
  figure_file('synthetic_data_generation.tmp.pdf', "2-dataset_characterisation/3-synthetic"),
  "synthetic_data_generation",
  "Workflow to generate the synthetic data. Starting from a state network, we extracted a regulatory network using real networks which is expected to induce such a set of state transitions when simulated. We simulated this network using a detailed model of gene regulation which includes the thermodynamics of transcription factor binding. Finally, we simulated the single-cell RNA-seq (scRNA-seq) experiment itself, by matching the distribution of expression and dropouts to a real reference dataset."
)
```

@@

```{r}
add_sfig(
  figure_file('expanded_ti_evaluation_overview_v1.tmp.pdf', 'evaluation_overview'),
  "extended_evaluation_overview",
  "Extended overview of our evaluation pipeline. From literature we extracted a quality control checklist, a list of trajectory inference methods, a set of real datasets containing a trajectory and real regulatory networks used to generate the synthetic data. We created a wrapper of each method, so that its output was transformed into a common probabilistic trajectory model, and used these to infer trajectories on all real and synthetic datasets using their default parameters. Using several similarity metrics, we compared the gold standard of the real and synthetic datasets with the inferred trajectories. We also evaluated the quality of each method using the quality control checklist, and used both the benchmark results and quality control results to produce a final set of guidelines for method's users.", 
  15, 
  7.5
)
```

@@

```{r}
add_sfig(
  figure_file(
    "real_synthetic_comparison.tmp.pdf",
    "5-optimise_parameters/5-real_synthetic_comparison"
  ), 
  "real_synthetic_comparison", 
  "Comparison between performance on real and synthetic datasets across trajectory types. Given in the top-left corner of each panel is the Pearson's correlation."
)
```

<!-- §4: Compare metrics -->
As the different metrics were selected to assess the correctness of a trajectory in various approaches, it is expected to observe differences in rankings of TI methods amongst the different metrics. While we saw no direct link between the edge flip scores versus the correlation or RF MSE metric, methods that obtained a high correlation score also tended to obtain higher RF MSE scores (`r ref("sfig", "metric_comparison")`). In addition, the methods that were able to detect more complex trajectory types also obtained higher edge flip scores, in comparison to methods whose trajectory topology was fixed to a simple trajectory topology. We will explore this issue in more detail in a further section.

```{r}
add_sfig(
  figure_file('metrics_comparison.tmp.pdf', "5-optimise_parameters/7-metric_comparison"), 
  "metric_comparison", 
  "Comparison between the different metrics across trajectory types. For each row, a different pairwise comparison is given between two metrics."
)
```

@@

```{r}
add_sfig(
  figure_file('trajtype_handle_comparison.tmp.pdf', "8-compare_topology"), 
  "trajtype_handle_comparison", 
  "Comparison between the performance of methods able to handle a trajectory type with those who don't. Shown is the performance on those datasets containing a particular trajectory type (top), according to several metrics (right)."
)
```

```{r}
add_sfig(
  figure_file('linear_split_comparison.tmp.pdf', "8-compare_topology"), 
  "linear_split_comparison", 
  "Comparison of the performance between linear and non-linear methods. Methods were split in linear methods (those which can either only detect cyclic or linear trajectories) and non-linear methods. Their performance is shown on linear and non-linear datasets according to several performance metrics (right). NS: Not significant, *: p-value < 0.05, **: p-value < 0.001, ***: p-value < 0.0001"
)
```

@@

```{r}
add_sfig(
  figure_file('topology_sensitivity.tmp.pdf', "8-compare_topology"), 
  "topology_sensitivity", 
  "Sensitivity of each TI method in predicting the correct trajectory topology. Shown are the % of datasets where a method is able to predict the exact correct trajectory topology, depending on the trajectory type present in the data. When a method is able to predict a particular trajectory type, the box is surrounded by a black border."
)
```

@@

```{r}
add_sfig(
  figure_file('required_priors_performance_comparison.tmp.pdf', "10-prior_information"), 
  "required_priors_performance_comparison", 
  "Comparison of method performance depending on whether the method required a certain prior or not. The top performing method which requires a particular type of prior information is labelled."
)
```

@@

```{r}
add_sfig(
  figure_file("method_components_scores.tmp.pdf","6-developer_guidelines"), 
  "method_components_scores", 
  "Score of methods containing particular algorithmic components. Shown are the algorithmic components present in four or more methods, and whether the peformance of methods containing a certain component was significantly higher than all the other methods. NS: Not significant. *: corrected p-value < 0.05 "
)
```

```{r}
add_stable(
  read_rds(figure_file(
    "component_importance.rds",
    "6-developer_guidelines"
  ))[[params$table_format]], 
  "component_importance", 
  "Importance values for algorithmic components. The overall performance was compared between methods containing a particular algorithmic component, and we calculated both a corrected p-value using a two-tailed Mann–Whitney U test and the increase in node purity importance measure using random forest classification."
)
```

@@

```{r}
add_sfig(
  figure_file(
    "qc_performance_comparison.tmp.pdf",
    "5-optimise_parameters/6-qc_performance_comparison"
  ), 
  "qc_performance_comparison", 
  "Comparison between method quality control scores and overall method performance."
)
```

@@

Our workflow to generate synthetic data is based on the well established workflow used in the evaluation of network inference methods [@schaffter_genenetweaversilicobenchmark_2011; @marbach_wisdom_2012] and consists of four main steps: network generation, simulation, gold standard extraction and simulation of the scRNA-seq experiment (`r ref("sfig", "synthetic_data_generation")`). At every step, we took great care to mimic real cellular regulatory networks as best as possible, while keeping the model simple and easily extendable. For every synthetic dataset, we used a random real dataset as a reference dataset (from those described earlier), making sure the number of variable genes and cells were similar.

@@

One of the main processes involved in cellular dynamic processes is gene regulation, where regulatory cascades and feedback loops lead to progressive changes in expression and decision making. The exact way a cell choses a certain path during its differentiation is still an active research field, although certain models have already emerged and been tested in vivo. One driver of bifurcation seems to be mutual antagonism, where genes [@xu_regulationbifurcatingcell_2015] strongly repress each other, forcing one of the two to become inactive [@graf_forcingcellschange_2009]. Such mutual antagonism can be modelled and simulated [@wang_quantifyingwaddingtonlandscape_2011; @ferrell_bistabilitybifurcationswaddington_2012]. Although such a two-gene model is simple and elegant, the reality is frequently more complex, with multiple genes (grouped into models) repressing each other [@yosef_dynamicregulatorynetwork_2013].

To simulate certain trajectory topologies, we therefore designed module networks in which the cells follow a particular trajectory topology given certain parameters (`r ref("sfig", "modulenets")`). Two module networks generated linear trajectories (linear and linear long), two generated simple forks (bifurcating and converging), one generated a complex fork (trifurcating), one generated a rooted tree (consecutive bifurcating) and two generated simple undirected graphs (bifurcating loop and bifurcating convergence).

From these module networks we generated gene regulatory networks in two steps: the main regulatory network was first generated, and extra target genes from real regulatory networks  were added. For each dataset, we used the same number of genes as were differentially expressed in the real datasets. 5% of the genes were assigned to be part of the main regulatory network, and were randomly distributed among all modules (with at least one gene per module). We sampled edges between these individual genes (according to the module network) using a uniform distribution between 1 and the number of possible targets in each module. To add additional target genes to the network, we assigned every regulator from the network to a real regulator in a real network (from regulatory circuits [@marbach_tissuespecificregulatorycircuits_2016]), and extracted for every regulator a local network around it using personalized pagerank (with damping factor set to 0.1), as implemented in the `page_rank` function of the *igraph* package.

```{r}
add_sfig(
  figure_file("modulenets.tmp.pdf", experiment_id = "2-dataset_characterisation/3-synthetic"),
  "modulenets", 
  "Module networks, milestone networks and the module dynamics for each of the different types of synthetic data", 
  15, 
  5
)
```

### Simulation of gene regulatory systems using thermodynamic models

To simulate the gene regulatory network, we used a system of differential equations similar to those used in evaluations of gene regulatory network inference methods [@marbach_wisdom_2012]. In this model, the changes in gene expression ($x_i$) and protein expression ($y_i$) are modeled using ordinary differential equations [@schaffter_genenetweaversilicobenchmark_2011] (ODEs):

@@

Because each cellular simulation follows the trajectory at its own speed, knowing the exact position of a cell within the trajectory topology is not straightforward. Furthermore, the speed at  which simulated cells make a decision between two or more alternative paths is highly variable. To estimate a cell's position during a simulation within the trajectory topology, we therefore used the known progression of the modules, given in `r ref("sfig", "modulenets", "c")`,  as a backbone. We smoothed the expression in each simulation using a rolling mean with a window of 50 time steps, and then calculated the average module expression along the simulation. We used dynamic time warping, implemented in the dtw R package [@giorgino_computingvisualizingdynamic_2009; @tormene_matchingincompletetime_2009], with an open end to align a simulation to all possible module progressions, and then picked the alignment which minimised the normalised distance between the simulation and the backbone. In case of cyclical trajectory topologies, the number of possible milestones a backbone could progress through was limited to 20.

## Expression normalisation pipeline
We used a standard single-cell RNA-seq preprocessing pipeline which applies parts of the scran and scater Bioconductor packages [@lun_stepbystepworkflowlowlevel_2016]. The advantages of this pipeline is that it works both with and without spike-ins, and includes a harsh cell filtering which looks at abnormalities in library sizes, mitochondrial gene expression, and number of genes expressed using median absolute deviations (set to 3). We required that a gene was expressed in at least 5% of the cells, and that it should have an average expression higher than 0.05. Furthermore, we used the pipeline to select the most highly variable genes, using a false discovery rate of 5% and a biological component higher than 0.5. As a final filter, we removed both zero genes and cells until convergence.

## Evaluation metrics
The importance of using multiple metrics to compare complex models has been stated repeatedly [@norel_selfassessmenttrap_2011]. We defined three metrics for comparing the likeness of predicted trajectories to a gold standard (`r ref("fig", "evaluation_overview", "c")`). Each metric assesses the performance of a different aspect of the trajectory (`r ref("sfig", "score_aspects")`); (i) the correlation metric measures the similarity in pairwise cell-cell distances; (ii) the RF MSE metric assesses whether cell neighbourhoods are similar in both trajectories; and (iii) the edge flip scores assesses similarity in trajectory topologies.

```{r}
add_sfig(
  figure_file("score_aspects.tmp.pdf", experiment_id="3-score_rules/5-score_aspects"), 
  "score_aspects", 
  "Different scores assess different aspects of the correspondence between a prediction and gold standard trajectory.",
  15, 
  8
)
```

### Correlation between geodesic distances
The similarity in cell ordering between two trajectories is assessed by calculating the geodesic distances between each pair of cells for both trajectories. The definition of a geodesic distance between two cells part of the common trajectory model will be demonstrated using a toy example (`r ref("sfig", "metric_geodesic")`).

```{r}
add_sfig(
  figure_file("metric_geodesic.tmp.pdf", experiment_id="evaluation_overview"), 
  "metric_geodesic", 
  "**The modified geodesic distances demonstrated on a toy example.** a) A toy example containing four milestones (W to Z) and five cells (a to e). b) The corresponding milestone network, milestone percentages and regions of delayed commitment, when the toy trajectory is converted to the common trajectory model. c) The calculations made for calculating the pairwise geodesic distances. d) A heatmap representation of the pairwise geodesic distances."
)
```

The geodesic distance between two cells depends on whether they are (i) on the same transition, or (ii) in different transitions. In the first case, the distance is defined as the product of the difference in milestone percentages and the length of the transition they both reside on. For cells $a$ and $b$ in the example, $d(a, b)$ is equal to $1 \times (0.9 - 0.2) = 0.7$. In the latter case, the distances between the cells and all of their neighbouring milestones will be calculated. These distances in combination with the milestone network are used to calculate the shortest path distance between the two cells. For cells $a$ and $c$ in the example, $d(a, X) = 1 \times 0.9$ and $d(c, X) = 3 \times 0.2$, and therefore $d(a, c) = 1 \times 0.9 + 3 \times 0.2$.

According to the defined common trajectory model (`r ref("fig", "evaluation_overview", "b")`), cells are also allowed to have a delayed commitment. In a region of delayed commitment, one milestone will be connected to all other milestones as per the milestone network. In this case, the distance between two cells both inside a region of delayed commitment is calculated as the manhattan distances between the milestone percentages weighted by the transition weights from the milestone network. For cells $d$ and $e$ in the example, $d(d, e)$ is equal to $0 \times (0.3 - 0.2) + 2 \times (0.7 - 0.2) + 3 \times(0.4 - 0.1)$, which is equal to $1.9$. The distance between two cells where one is part of a region of delayed commitment is calculated similarly to the previous paragraph, by first calculating the between the cells and their neighbouring milestones first, then calculating the shortest path distances between the two.

Finally, calculating all pairwise distances between cells would scale poorly for trajectories with large numbers of cells. For this reason, a set of waypoint cells are defined *a priori*, and only the distances between the waypoint cells and all other cells is calculated, in order to calculate the correlation of geodesic distances of two trajectories. The waypoints are determined by viewing each milestone, transition and region of delayed commitment as a collection of cells, and sampling cells from the different collections weighted by the total number of cells within that collection. For calculating the correlation of geodesic distances between two trajectories, the distances between all cells and the union of both waypoint sets is computed. For the benchmark evaluation, the total number of waypoints sampled from a trajectory was one hundred.

@@

Although the correlation between geodesic distances directly assesses the position of the cells in the trajectory, a bad correlation does not directly imply that similar cells were not grouped together by the method, as illustrated in `r ref("sfig", "score_aspects", c("b", "c"))`. For example, certain methods will inevitably reach an incorrect ordering because they cannot handle the correct trajectory type, but these methods could still correctly place similar cells next to each other. We therefore also included a metric which looks at the local neighbourhood of each cell, and assesses whether this neighbourhood can accurately predict the position of this cell in the gold standard. We used a Random Forest regression, implemented in the *ranger* package [@wright_rangerfastimplementation_2017] to separately predict milestone percentages of every cell in the gold standard, using the milestone percentages of these cells in the prediction as features. We then used the out-of-bag mean-squared error on these percentages to score each method's capability of predicting the correct neighbourhood of each cell.

@@

As a third independent score, we assessed the similarity between the milestone network topologies. We first simplified each network, by merging consecutive linear edges into one edge, and adding new milestones within self loops such that $A \rightarrow A$ would be converted $A \rightarrow B \rightarrow C \rightarrow D$, by adding an intermediate node to linear networks. Because we are interested in the overall similarity between two topologies irrespective of the direction of the edges, the network was made undirected. Next, we define the edge flip score as the minimal number of edges which should be added or removed to convert one network into the other, divided by the total number of edges in both networks. This problem is equivalent to the maximum common edge subgraph problem, a known NP-hard problem without a scalable solution [@bahiense_maximumcommonedge_2012]. We implemented a branch and bound approach for this problem, by first enumerating all possible edge additions and removals with the minimal number of edges (the edge difference between the two networks) and if none of the new networks was isomorphic, we tried out all solutions with additional two edge changes. To further limit the search space, we made sure the degree distributions between the two networks were  similar, before assessing whether the two networks were isomorphic using the BLISS algorithm [@junttila_engineeringefficientcanonical_2007], as implemented in the R *igraph* package.

A comparison of edge flip scores between common trajectory topologies is illustrated in `r ref("fig", "edge_flip_comparison")`.

```{r}
add_sfig(
  figure_file("edge_flip_comparison.tmp.pdf", experiment_id="3-score_rules/4-edge_flip"), 
  "edge_flip_comparison", 
  paste0("Edge flip score values for common trajectory topologies, defined in ", ref("sfig", "modulenets"), "."), 
  15, 
  5
)
```

### Aggregation of scores
`r ref("fig", "aggregation_of_results")` illustrates the full set of aggregations performed on the raw scores in order to arrive at the final ranking of methods. In order to make the scores produced by the different metrics comparable to one another, across datasets of varying difficulty, the raw scores are percentile rank transformed per metric per dataset to a $[0,1]$ range. The arithmetic mean is calculated across replicates. In order to ensure a method only obtains a high score if it scores well on all three metrics, the harmonic mean across metrics is calculated. At this point, the aggregated scores of all methods across all datasets is combined. As there can be an overrepresentation of datasets of a certain trajectory type, first an arithmetic mean is calculated per trajectory type, followed by an overall arithmetic mean across all trajectory types, thus obtaining a ranking of the methods.

```{r}
add_sfig(
  figure_file("metric_aggregation_overview_v1.tmp.pdf", experiment_id="evaluation_overview"), 
  "aggregation_of_results", 
  "Aggregation methodology of metric scores.", 
  15, 
  5
)
```

@@

### Method execution
Each execution of a method on a dataset was performed in a separate task as part of a gridengine job. Each task was allocated one CPU core of an Intel(R) Xeon(R) CPU E5-2665 0 @ 2.40GHz, and was provided a maximum of 32GB in memory and 6 hours of wall time. One R session was started for each task, with the environment variable `R_MAX_NUM_DLLS` set to 500. The `base::set.seed` was overridden in order to prevent stochastic TI methods from pretending to be deterministic and/or robust. Timings of methods were measured for different steps along the executions, including preprocessing, postprocessing, each of the different metrics, and the method itself.

### Effect of prior information
To assess the effect of prior information on the performance of a method, we compared the performance of methods which do or do not require the prior information using a two-tailed Mann–Whitney U test. P-values were controlled for multiple testing using Benjamini-Hochberg correction.
